---
title: "lab_01"
author: "Derek Willis"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

```{r}
glimpse (umd_courses)
umd_courses |> 
  group_by (id, instructors) |> 
  summarize (
    count_classes = n()
  )
```


## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you?
Engineering is the top, followed by first-year innovation and university courses. It's interesting that there's a huge gap between #1 and #2 - roughly 100 seats.

```{r}
umd_courses |> 
  group_by(department) |> 
  summarize (
    average_seats = mean(seats)
  ) |> 
  arrange(desc(average_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first. 


```{r}
umd_courses |> 
  group_by(id) |> 
  summarize (
    count_classes = n()
    ) |> 
arrange(desc(count_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? **A3**
Unless you're including NA or Instructor: TBA, 73 is the max. Elsa Barkley Brown is the instructor who's taught that many classes.

```{r}
umd_courses_fin <- umd_courses |> 
  group_by (instructors) |> 
  summarize (
    count_classes = n()
    ) |> 
  arrange(desc(count_classes))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**

It would've been easiest to use a GUI like Excel instead of typing in code with R. Or, failing that, some kind of a shortcut to make finding the mean easier. I'm pretty confident that I got the right answer, but I'm still not totally clear on how the count() function works. How does it know to filter out/count duplicates?

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**

Question: How many classes has the Jewish Studies department had, and are the number of seats available this semester more or less than average? I need to filter for Jewish studies and then sum the number of classes for each semester aggregated by semester. Below is my attempt.

```{r}
umd_courses_fin |> 
  group_by (count_classes)
```
```{r}
```{r}
umd_courses |> 
  group_by (department, id, instructors) |> 
  filter(department == "Jewish Studies") |> 
  summarize (
    count_classes = n()
    ) |> 
  arrange(desc(count_classes)) |> 
  filter(!is.na(instructors))
```
```

